Parametricity for Primitive Nested Types and GADTs

### 2021-07-06 21-lmcs/21-lmcs.tex
6th  LMCS revisions
LMCS revisions 6

\usepackage{bussproofs}
\EnableBpAbbreviations

@@

In~\cite{pit00}, Pitts adds list ADTs to full System F with a
term-level fixpoint primitive. Other ADTs are included
in~\cite{pit98}, but nested types are not expressible in either
syntax. Pitts constructs parametric models for his calculi based on
operational, rather than categorical, semantics. A benefit of using
operational semantics to build parametric models is that it avoids
needing to work in a suitable metatheory to accommodate System F's
impredicativity. It is well-known that there are no set-based
parametric models of System F~\cite{rey84}, so parametric models for
it and its extensions are often constructed in a syntactic metatheory
such as the impredicative Calculus of Inductive Constructions (iCIC).
By adding primitive nested types to a Hindley-Milner-style calculus
and working in a categorical setting we side-step such metatheoretic
distractions. It is important to note that different consequences of
parametricity are available in syntactic and semantic
metatheories. Consequences of parametricity are possible for both
closed and open System F terms in a syntactic metatheory --- although
not all that can be formulated can be always proved; see, e.g., the
end of Section~7 of~\cite{bm98}. By contrast, in a categorical
metatheory consequences of parametricity are expressible only for {\em
  closed} terms. For this reason, validating the standard consequences
of parametricity for closed terms is --- going all the way back to
Reynolds~\cite{rey83} --- all that is required for a model of
parametricity to be considered good.

@@

\vspace*{-0.2in}

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[3ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\BIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$}  
\DisplayProof
\\[4ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
\\[3ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof
\end{array}\]
\end{defi}
We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
Definition~\ref{def:wftypes} ensures that the expected weakening rules
for well-formed types hold, i.e., if $\Gamma; \Phi \vdash F$ is
well-formed, then both $\Gamma, \ol{\psi}; \Phi \vdash F$ and $\Gamma;
\Phi, \ol{\phi} \vdash F$ are also well-formed. Note, however, that
weakening does not change the contexts in which types can be formed.
For example, the functorial context in which a type $\Nat^\emptyset F
\,G$ is formed must be empty, although it can be weakened by a
functorial variable $\beta$ {\em after} the type is formed; by
contrast, we cannot {\em first} weaken the functorial contexts of $F$
and $G$ to $\Gamma; \beta \vdash F$ and $\Gamma; \beta \vdash G$, and
{\em then} form $\Nat^\emptyset F \,G$ in the weakened context
$\Gamma; \beta$.  If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset
\vdash G$, then our rules allow formation of the type
$\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$, which represents the
System F type $\Gamma \vdash F \to G$. Similarly, if
$\Gamma;\ol{\alpha} \vdash F$, then our rules allow formation of the
type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which
represents the System F type $\Gamma; \emptyset \vdash \forall
\ol\alpha . F$. However, some System F types, such as $\forall
\alpha. (\alpha \to \alpha) \to \alpha$, are not representable in our
calculus. Note that, in the rule for $\mu$-types, no variables in
$\Phi$ appear in the body $F$ of $\mu \phi. \lambda \ol\alpha.F$. This
will be critical to proving the Identity Extension Lemma for our
calculus.

@@

\begin{defi}\label{def:set-sem}
The {\em set interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$
is defined by:
\begin{align*}
  \setsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \setsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \setsem{\Gamma; \emptyset
    \vdash \Nat^{\ol{\alpha}}
    \,F\,G}\rho &= \{\eta : \lambda \ol{A}. \,\setsem{\Gamma;
    \ol{\alpha} \vdash
    F}\rho[\ol{\alpha := A}] 
      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; 
        \ol{\alpha} \vdash G}\rho[\ol{\alpha := A}] \\ 
      &\hspace{0.3in}|~\forall \overline{A}, \overline{B} :
      \set. \forall \overline{R : \rel(A, B)}.\\ 
      &\hspace{0.4in}(\eta_{\overline{A}}, \eta_{\overline{B}})
      : \relsem{\Gamma; \ol{\alpha} \vdash F}\Eq_{\rho}[\ol{\alpha := R}]
      \rightarrow \relsem{\Gamma; \ol{\alpha} \vdash
        G}\Eq_{\rho}[\ol{\alpha := R}] \} \\
  \setsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
  (\rho\phi)\,\ol{\setsem{\Gamma;\Phi \vdash
    F}\rho}\\
  \setsem{\Gamma;\Phi \vdash F+G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho +
  \setsem{\Gamma;\Phi \vdash G}\rho\\
  \setsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho \times
  \setsem{\Gamma;\Phi \vdash G}\rho\\ 
  \setsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}\rho}\\
    \text{where } T^\set_{H,\rho}\,F & = \lambda
  \ol{A}. \setsem{\Gamma;\phi, \ol{\alpha} \vdash
    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
  \text{and } T^\set_{H,\rho}\,\eta &= \lambda
  \ol{A}. \setsem{\Gamma;\phi, \ol{\alpha} \vdash
    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}]
\end{align*}
\end{defi}
If $\rho \in \setenv$ and $\vdash F$ then we write $\setsem{\vdash F}$
instead of $\setsem{\vdash F}\rho$ since the environment is
immaterial. The third clause of Definition~\ref{def:set-sem} does
indeed define a set: local finite presentability of $\set$ and
$\omega$-cocontinuity of $\setsem{\Gamma;\ol{\alpha} \vdash F}$
ensure that $\{\eta : \lambda \ol{A}.\,\setsem{\Gamma;\ol{\alpha}
  \vdash F}\rho[\ol{\alpha := A}] \Rightarrow \lambda
\ol{A}.\,\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha := A}]\}$
(containing $\setsem{\Gamma;\emptyset \vdash
  \Nat^{\ol{\alpha}}\,F\,G}\rho$) can be embedded in the set
\[ \prod_{\substack{\ol{S} =
    (S_1,...,S_{|\ol{\alpha}|})\\ S_1,...,S_{|\ol{\alpha}|} \mbox{ are
      finite cardinals}}} \hspace*{-0.4in}(\setsem{\Gamma;\ol{\alpha}
  \vdash G}\rho[\ol{\alpha := S}])^{(\setsem{\Gamma;\ol{\alpha} \vdash
    F}\rho[\ol{\alpha := S}])}\] Also, $\setsem{\Gamma; \emptyset
  \vdash \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous since it is
constant on $\omega$-directed sets; indeed, it is constant on
$\omega$-directed sets because Definition~\ref{def:set-env} ensures
that restrictions to $\mathbb{T}^k$ of morphisms between set
environments are identities.  Interpretations of $\Nat$-types ensure
that $\setsem{\Gamma \vdash F \to G}$ and $\setsem{\Gamma \vdash
  \forall \ol\alpha. F}$ are as expected in any parametric model.

@@

The interpretations in Definitions~\ref{def:rel-sem}
and~\ref{def:rel-sem-funcs} below respect weakening, and also ensure
that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma \vdash
  \forall \ol \alpha. F}$ are as expected in any parametric model.  As
for set interpretations, $\relsem{\Gamma; \emptyset \vdash
  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous; indeed, it is
constant on $\omega$-directed sets because
Definition~\ref{def:reln-env} ensures that restrictions to
$\mathbb{T}^k$ of morphisms between relational environments are
identities. If $\rho \in \relenv$ and $\vdash F$, then we write
$\relsem{\vdash F}$ instead of $\relsem{\vdash F}\rho$.  For the last
clause in Definition~\ref{def:rel-sem} to be well-defined we need
$T_{H,\rho}$ to be an $\omega$-cocontinuous endofunctor on $RT$ so
that, by Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$
is defined in terms of $\relsem{\Gamma;\phi, \ol{\alpha} \vdash H}$,
this means that relational interpretations of types must be
$\omega$-cocontinuous functors from $\relenv$ to $RT_0$, which in turn
entails that the actions of relational interpretations of types on
objects and on morphisms in $\relenv$ are intertwined. As for set
interpretations, we know from~\cite{jp19} that, for every $\Gamma;
\ol{\alpha} \vdash F$, $\relsem{\Gamma; \ol{\alpha} \vdash F}$ is
actually in $[\rel^k,\rel]$ where $k = |\ol \alpha|$. We first define
the actions of each of these functors on morphisms between
environments, and then argue that they are well-defined and have the
required properties. To do this, we extend $T_H$ to a {\em functor}
from $\relenv$ to $[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an
object $\rho \in \relenv$ is given by the higher-order functor
$T_{H,\rho}$ whose actions on objects and morphisms are given in
Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
\to \rho'$ is the higher-order natural transformation $T_{H,f} :
T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\rel^k,\rel]$ is
the natural transformation $T_{H,f}\, F : T_{H,\rho}\, F \to
T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
F)_{\ol{R}} = \relsem{\Gamma; \phi,\ol{\alpha} \vdash H}f[\phi :=
  \id_F][\ol{\alpha := \id_R}]$.

@@

\begin{lem}\label{lem:graph-reln-functors}
If $F,G : [\set^k,\set]$, and if $\alpha : F \to G$ is a natural
transformation, then $\graph{\alpha}$ is in $RT_k$.
\end{lem}
\proof
Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
$\graph{\alpha}^* : [\rel^k,\rel]$. Let $\overline{R :
  \rel(A, B)}$, $\overline{S : \rel(C, D)}$, and $\overline{(\beta,
  \beta') : R \to S}$. We want to show that there exists a morphism
$\epsilon : \alpha^\wedge\overline{R} \to \alpha^\wedge\overline{S}$
such that the diagram on the left below commutes. Since
$\ol{(\beta,\beta') : R \to S}$, there exist $\overline{\gamma : R \to
  S}$ such that each diagram in the middle commutes.
Moreover, since both $h_{\overline{C \times D}} \circ
F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make the diagram
on the right commute, they must be equal.
%\begin{figure*}[ht]
%\vspace*{-0.1in}
%  \hspace*{-0.85in}
  \begin{minipage}[b]{0.25\linewidth}
 {\small    \[
    \begin{tikzcd}
        \alpha^\wedge\overline{R}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
        \ar[d, "{\epsilon}"']
        & F\overline{A} \times G\overline{B}
        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
        \alpha^\wedge\overline{S}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
        & F\overline{C} \times G\overline{D}
    \end{tikzcd}
    \]}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
{\small    \[
    \begin{tikzcd}
        R_i
        \ar[d, "{\gamma_i}"']
        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
        &A_i \times B_i
        \ar[d, "{\beta_i \times \beta'_i}"] \\
        S_i
        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
        &C_i \times D_i
    \end{tikzcd}
    \]}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
{\footnotesize \[
      \begin{tikzcd}[row sep = large]
          F\overline{C}
          &F\overline{C} \times F\overline{D}
          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
          &F\overline{D}
          \ar[r, "{\alpha_{\ol{D}}}"]
          &G\overline{D}\\
          &F(\overline{A \times B})
          \ar[u, dashed, "{\exists !}"]
          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
          \ar[urr, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \times \beta'})}"']
      \end{tikzcd}
      \]}
\end{minipage}
%\end{figure*}

@@

\noindent
We therefore get that the right-hand square in the diagram on the left
below commutes, and thus that the entire diagram does as well.
Finally, by the left-lifting property of $q_{F^\wedge\overline{R}}$
with respect to $\iota_{F^\wedge\overline{S}}$ given by the mono-epi
factorization system, there exists an $\epsilon$ such that the diagram
on the right below commutes as desired.
\begin{figure*}[ht]
  \vspace*{-0.1in}
  \hspace*{-0.5in}
  \begin{minipage}[b]{0.45\linewidth}
{\footnotesize \[
      \begin{tikzcd}
          F\overline{R}
          \ar[d, "{F\overline{\gamma}}"']
          \ar[r, hookrightarrow, "{F\overline{\iota_R}}"]
          \ar[rr, bend left, "{h_{\overline{R}}}"]
          &F(\overline{A \times B})
          \ar[d, "{F(\overline{\beta \times \beta'})}"]
          \ar[r, "{h_{\overline{A \times B}}}"]
          &F\overline{A} \times G\overline{B}
          \ar[d, "{F\overline{\beta} \times F\overline{\beta'}}"] \\
          F\overline{S}
          \ar[r, hookrightarrow, "{F\overline{\iota_S}}"']
          \ar[rr, bend right, "{h_{\overline{S}}}"']
          &F(\overline{C \times D})
          \ar[r, "{h_{\overline{C \times D}}}"']
          &F\overline{C} \times G\overline{D}
      \end{tikzcd}
      \]}
\end{minipage}
  \vspace*{-0.5in}
  \begin{minipage}[b]{0.45\linewidth}
      {\footnotesize
        \[  \vspace*{0.4in}
      \begin{tikzcd}
          F\overline{R}
          \ar[d, "{F\overline{\gamma}}"']
          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{R}}}"]
          &\alpha^\wedge\overline{R}
          \ar[d, dashed, "{\epsilon}"]
          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
          &F\overline{A} \times G\overline{B}
          \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
          F\overline{S}
          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{S}}}"']
          &\alpha^\wedge\overline{S}
          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
          &F\overline{C} \times G\overline{D}
      \end{tikzcd}
      \]}
\end{minipage}
\end{figure*}

>>>>

\usepackage{bussproofs}
\EnableBpAbbreviations
\def\ruleScoreFiller{\hrule height 0.8pt}

@@

In~\cite{pit00}, Pitts adds list ADTs to full System F with a
term-level fixpoint primitive. Other ADTs are included
in~\cite{pit98}, but nested types are not expressible in either
syntax. Pitts constructs parametric models for his calculi based on
operational, rather than categorical, semantics. A benefit of using
operational semantics to build parametric models is that it avoids
needing to work in a suitable metatheory to accommodate System F's
impredicativity. It is well-known that there are no set-based
parametric models of System F~\cite{rey84}, so parametric models for
it and its extensions are often constructed in a syntactic metatheory
such as the impredicative Calculus of Inductive Constructions (iCIC).
By adding primitive nested types to a Hindley-Milner-style calculus
and working in a categorical setting we side-step such metatheoretic
distractions.
%It is important to note that different consequences of
%parametricity are available in syntactic and semantic
%metatheories. Consequences of parametricity are possible for both
%closed and open System F terms in a syntactic metatheory --- although
%not all that can be formulated can be always proved; see, e.g., the
%end of Section~7 of~\cite{bm98}. By contrast, in a semantic
%metatheory consequences of parametricity are expressible only for {\em
%  closed} terms. For this reason, validating the standard consequences
%of parametricity for closed terms is --- going all the way back to
%Reynolds~\cite{rey83} --- all that is required for a model of
%parametricity to be considered good.

@@

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]\\[3ex]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[5ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[5ex]
\AXC{$\Gamma;\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\BIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$}  
\DisplayProof
\\[6ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
\\[5ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof\\[2ex]
\end{array}\]
\end{defi}
We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
Definition~\ref{def:wftypes} ensures that the expected weakening rules
for well-formed types hold, i.e., if $\Gamma; \Phi \vdash F$ is
well-formed, then both $\Gamma, \ol{\psi}; \Phi \vdash F$ and $\Gamma;
\Phi, \ol{\phi} \vdash F$ are also well-formed. Note, however, that
weakening does not change the contexts in which types can be formed.
For example, the functorial context in which a type $\Nat^\emptyset F
\,G$ is formed must be empty, although it can be weakened by a
functorial variable $\beta$ {\em after} the type is formed; by
contrast, we cannot {\em first} weaken the functorial contexts of $F$
and $G$ to $\Gamma; \beta \vdash F$ and $\Gamma; \beta \vdash G$, and
{\em then} form $\Nat^\emptyset F \,G$ in the weakened context
$\Gamma; \beta$.  If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset
\vdash G$, then our rules allow formation of the type
$\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$, which represents the
System F type $\Gamma \vdash F \to G$. Similarly, if
$\Gamma;\ol{\alpha} \vdash F$, then our rules allow formation of the
type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which
represents the System F type $\Gamma; \emptyset \vdash \forall
\ol\alpha . F$. However, some System F types, such as $\forall
\alpha. (\alpha \to \alpha) \to \alpha$, are not representable in our
calculus. Note that, in the rule for $\mu$-types, no variables in
$\Phi$ appear in the body $F$ of $\mu \phi. \lambda \ol\alpha.F$. This
will be critical to proving the Identity Extension Lemma for our
calculus.

@@

\begin{defi}\label{def:set-sem}
The {\em set interpretation} $\setsem{\cdot} : \F \to [\setenv, \set]$
is defined by:
\begin{align*}
  \setsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \setsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \setsem{\Gamma; \emptyset
    \vdash \Nat^{\ol{\alpha}}
    \,F\,G}\rho &= \{\eta : \lambda \ol{A}. \,\setsem{\Gamma;
    \ol{\alpha} \vdash
    F}\rho[\ol{\alpha := A}] 
      \Rightarrow \lambda \ol{A}.\,\setsem{\Gamma; 
        \ol{\alpha} \vdash G}\rho[\ol{\alpha := A}] \\ 
      &\hspace{0.3in}|~\forall \overline{A}, \overline{B} :
      \set. \forall \overline{R : \rel(A, B)}.\\ 
      &\hspace{0.4in}(\eta_{\overline{A}}, \eta_{\overline{B}})
      : \relsem{\Gamma; \ol{\alpha} \vdash F}\Eq_{\rho}[\ol{\alpha := R}]
      \rightarrow \relsem{\Gamma; \ol{\alpha} \vdash
        G}\Eq_{\rho}[\ol{\alpha := R}] \} \\
  \setsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
  (\rho\phi)\,\ol{\setsem{\Gamma;\Phi \vdash
    F}\rho}\\
  \setsem{\Gamma;\Phi \vdash F+G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho +
  \setsem{\Gamma;\Phi \vdash G}\rho\\
  \setsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \setsem{\Gamma;\Phi \vdash F}\rho \times
  \setsem{\Gamma;\Phi \vdash G}\rho\\ 
  \setsem{\Gamma;\Phi \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
    T^\set_{H,\rho})\ol{\setsem{\Gamma;\Phi \vdash G}\rho}\\
    \text{where } T^\set_{H,\rho}\,F & = \lambda
  \ol{A}. \setsem{\Gamma;\phi, \ol{\alpha} \vdash
    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
  \text{and } T^\set_{H,\rho}\,\eta &= \lambda
  \ol{A}. \setsem{\Gamma;\phi, \ol{\alpha} \vdash
    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}]
\end{align*}
\end{defi}
If $\rho \in \setenv$ and $\vdash F$ then we write $\setsem{\vdash F}$
instead of $\setsem{\vdash F}\rho$ since the environment is
immaterial. The third clause of Definition~\ref{def:set-sem} does
indeed define a set: local finite presentability of $\set$ and
$\omega$-cocontinuity of $\setsem{\Gamma;\ol{\alpha} \vdash F}$
ensure that $\{\eta : \lambda \ol{A}.\,\setsem{\Gamma;\ol{\alpha}
  \vdash F}\rho[\ol{\alpha := A}] \Rightarrow \lambda
\ol{A}.\,\setsem{\Gamma;\ol{\alpha} \vdash G}\rho[\ol{\alpha := A}]\}$
(containing $\setsem{\Gamma;\emptyset \vdash
  \Nat^{\ol{\alpha}}\,F\,G}\rho$) can be embedded in the set
\[ \prod_{\substack{\ol{S} =
    (S_1,...,S_{|\ol{\alpha}|})\\ S_1,...,S_{|\ol{\alpha}|} \mbox{ are
      finite cardinals}}} \hspace*{-0.4in}(\setsem{\Gamma;\ol{\alpha}
  \vdash G}\rho[\ol{\alpha := S}])^{(\setsem{\Gamma;\ol{\alpha} \vdash
    F}\rho[\ol{\alpha := S}])}\] Also, $\setsem{\Gamma; \emptyset
  \vdash \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous since it is
constant (in particular, on $\omega$-directed sets). This is because
Definition~\ref{def:set-env} ensures that restrictions to
$\mathbb{T}^k$ of morphisms between set environments are identities.
Interpretations of $\Nat$-types ensure that $\setsem{\Gamma \vdash F
  \to G}$ and $\setsem{\Gamma \vdash \forall \ol\alpha. F}$ are as
expected in any parametric model.

@@

The interpretations in Definitions~\ref{def:rel-sem}
and~\ref{def:rel-sem-funcs} below respect weakening, and also ensure
that $\relsem{\Gamma \vdash F \to G}$ and $\relsem{\Gamma \vdash
  \forall \ol \alpha. F}$ are as expected in any parametric model.  As
for set interpretations, $\relsem{\Gamma; \emptyset \vdash
  \Nat^{\ol\alpha} F\,G}$ is $\omega$-cocontinuous; indeed, it is
constant (on $\omega$-directed sets) because
Definition~\ref{def:reln-env} ensures that restrictions to
$\mathbb{T}^k$ of morphisms between relational environments are
identities. If $\rho \in \relenv$ and $\vdash F$, then we write
$\relsem{\vdash F}$ instead of $\relsem{\vdash F}\rho$.  For the last
clause in Definition~\ref{def:rel-sem} to be well-defined we need
$T_{H,\rho}$ to be an $\omega$-cocontinuous endofunctor on $RT$ so
that, by Lemma~\ref{lem:fp}, it admits a fixpoint. Since $T_{H,\rho}$
is defined in terms of $\relsem{\Gamma;\phi, \ol{\alpha} \vdash H}$,
this means that relational interpretations of types must be
$\omega$-cocontinuous functors from $\relenv$ to $RT_0$, which in turn
entails that the actions of relational interpretations of types on
objects and on morphisms in $\relenv$ are intertwined. As for set
interpretations, we know from~\cite{jp19} that, for every $\Gamma;
\ol{\alpha} \vdash F$, $\relsem{\Gamma; \ol{\alpha} \vdash F}$ is
actually in $[\rel^k,\rel]$ where $k = |\ol \alpha|$. We first define
the actions of each of these functors on morphisms between
environments, and then argue that they are well-defined and have the
required properties. To do this, we extend $T_H$ to a {\em functor}
from $\relenv$ to $[[\rel^k,\rel],[\rel^k,\rel]]$. Its action on an
object $\rho \in \relenv$ is given by the higher-order functor
$T_{H,\rho}$ whose actions on objects and morphisms are given in
Definition~\ref{def:rel-sem-funcs}. Its action on a morphism $f : \rho
\to \rho'$ is the higher-order natural transformation $T_{H,f} :
T_{H,\rho} \to T_{H,\rho'}$ whose action on any $F : [\rel^k,\rel]$ is
the natural transformation $T_{H,f}\, F : T_{H,\rho}\, F \to
T_{H,\rho'}\, F$ whose component at $\ol{R}$ is $(T_{H,f}\,
F)_{\ol{R}} = \relsem{\Gamma; \phi,\ol{\alpha} \vdash H}f[\phi :=
  \id_F][\ol{\alpha := \id_R}]$.

@@

\begin{lem}\label{lem:graph-reln-functors}
If $F,G : [\set^k,\set]$, and if $\alpha : F \to G$ is a natural
transformation, then $\graph{\alpha}$ is in $RT_k$.
\end{lem}
\proof
Clearly, $\graph{\alpha}^*$ is $\omega$-cocontinuous, so
$\graph{\alpha}^* : [\rel^k,\rel]$. Let $\overline{R :
  \rel(A, B)}$, $\overline{S : \rel(C, D)}$, and $\overline{(\beta,
  \beta') : R \to S}$. We want to show that there exists a morphism
$\epsilon : \alpha^\wedge\overline{R} \to \alpha^\wedge\overline{S}$
such that the diagram on the left below commutes. Since
$\ol{(\beta,\beta') : R \to S}$, there exist $\overline{\gamma : R \to
  S}$ such that each diagram in the middle commutes.
Moreover, since both $h_{\overline{C \times D}} \circ
F(\overline{\beta \times \beta'})$ and $(F\overline{\beta} \times
G\overline{\beta'}) \circ h_{\overline{A \times B}}$ make the diagram
on the right commute, they must be equal.
\begin{figure*}[ht]
%\vspace*{-0.1in}
  \hspace*{-1in}
  \begin{minipage}[b]{0.25\linewidth}
 {\small    \[
    \begin{tikzcd}
        \alpha^\wedge\overline{R}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
        \ar[d, "{\epsilon}"']
        & F\overline{A} \times G\overline{B}
        \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
        \alpha^\wedge\overline{S}
        \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
        & F\overline{C} \times G\overline{D}
    \end{tikzcd}
    \]}
\end{minipage}\hspace*{0.2in}
\begin{minipage}[b]{0.25\linewidth}
{\small    \[
    \begin{tikzcd}
        R_i
        \ar[d, "{\gamma_i}"']
        \ar[r, hookrightarrow, "{\iota_{R_i}}"]
        &A_i \times B_i
        \ar[d, "{\beta_i \times \beta'_i}"] \\
        S_i
        \ar[r, hookrightarrow, "{\iota_{S_i}}"]
        &C_i \times D_i
    \end{tikzcd}
    \]}
\end{minipage}\hspace*{0.2in}
\begin{minipage}[b]{0.25\linewidth}
{\footnotesize \[
      \begin{tikzcd}[row sep = large]
          F\overline{C}
          &F\overline{C} \times G\overline{D}
          \ar[l, "{\pi_1}"'] \ar[r, "{\pi_2}"]
          &G\overline{D}\\
%          \ar[r, "{\alpha_{\ol{D}}}"]
%          &G\overline{D}\\
          &F(\overline{A \times B})
          \ar[u, dashed, "{\exists !}"]
          \ar[ul, "{F\ol{\pi_1} \circ F(\overline{\beta \times \beta'})}"]
          \ar[ur, "{\alpha_{\ol{D}} \circ F\ol{\pi_2} \circ F(\overline{\beta \times \beta'})}"']
      \end{tikzcd}
      \]}
\end{minipage}
\end{figure*}

@@

\noindent
We therefore get that the right-hand square in the diagram on the left
below commutes, and thus that the entire diagram does as well.
Finally, by the left-lifting property of $q_{F^\wedge\overline{R}}$
with respect to $\iota_{F^\wedge\overline{S}}$ given by the mono-epi
factorization system, there exists an $\epsilon$ such that the diagram
on the right below commutes as desired.
\begin{figure*}[ht]
  \vspace*{-0.1in}
  \hspace*{-0.5in}
  \begin{minipage}[b]{0.45\linewidth}
{\footnotesize \[
      \begin{tikzcd}
          F\overline{R}
          \ar[d, "{F\overline{\gamma}}"']
          \ar[r, hookrightarrow, "{F\overline{\iota_R}}"]
          \ar[rr, bend left, "{h_{\overline{R}}}"]
          &F(\overline{A \times B})
          \ar[d, "{F(\overline{\beta \times \beta'})}"]
          \ar[r, "{h_{\overline{A \times B}}}"]
          &F\overline{A} \times G\overline{B}
          \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
          F\overline{S}
          \ar[r, hookrightarrow, "{F\overline{\iota_S}}"']
          \ar[rr, bend right, "{h_{\overline{S}}}"']
          &F(\overline{C \times D})
          \ar[r, "{h_{\overline{C \times D}}}"']
          &F\overline{C} \times G\overline{D}
      \end{tikzcd}
      \]}
\end{minipage}
  \vspace*{-0.5in}
  \begin{minipage}[b]{0.45\linewidth}
      {\footnotesize
        \[  \vspace*{0.4in}
      \begin{tikzcd}
          F\overline{R}
          \ar[d, "{F\overline{\gamma}}"']
          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{R}}}"]
          &\alpha^\wedge\overline{R}
          \ar[d, dashed, "{\epsilon}"]
          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{R}}}"]
          &F\overline{A} \times G\overline{B}
          \ar[d, "{F\overline{\beta} \times G\overline{\beta'}}"] \\
          F\overline{S}
          \ar[r, twoheadrightarrow, "{q_{\alpha^\wedge\overline{S}}}"']
          &\alpha^\wedge\overline{S}
          \ar[r, hookrightarrow, "{\iota_{\alpha^\wedge\overline{S}}}"']
          &F\overline{C} \times G\overline{D}
      \end{tikzcd}
      \]}
\end{minipage}
\end{figure*}
